The Poisson approximate likelihood compared to the particle filter

### 2024-07-26 ms.Rnw
enable cond-lik plot

<<OvOv-model>>=
# measurement model 
dmeas <- Csnippet("
                  if (ISNA(cases1)) {
                  lik = (give_log) ? 0 : 1;
                  } else {
                        lik =  dbinom(cases1, H1, q1, 1) +
                        dbinom(cases2, H2, q2, 1) +
                        dbinom(cases3, H3, q3, 1);

@@

<<maximization1>>=
set.seed(123)
sir_box <- list(
  beta1=c(10,15),
  beta2=c(0.2,0.4),
  beta3=c(0.3,0.5),
  phi=c(0.01,0.3),
  beta11=c(0.1,0.2),
  sigma_q=c(0.001,0.1),
  sigma_xi=c(65,70),
  S10 = c(0.01,0.05),
  I10 = c(0.0001, 0.0005),
  R10 = c(0.01, 0.05),
  S20 = c(0.01,0.7),
  I20 = c(0.000001,0.00002),
  R20 = c(0.001,0.005),
  S30 = c(0.01,0.5),
  I30 = c(0.000001, 0.0001),
  R30 = c(0.0001,0.005),
  #fixed params
  gamma = c(1,1),
  delta1=c(1/(5*52),1/(5*52)),
  delta2=c(1/(55*52),1/(55*52)),
  alpha=c(1/(78.86912*52),1/(78.86912*52)),
  mu=c(0,0),
  N=c(82372825,82372825),
  omega=c(1/(1*52),1/(1*52))
)

@@

<<maximization2>>=
set.seed(1234)
sd_02 = 2/3

@@

<<maximization3>>=
set.seed(12345)
sd_03           = (2/3)^2

@@

<<pf-sim-100>>=
# Set WWR's MLE
wwr_mle_raw <- c(params_fixed, 
             "beta1" = 11.48,
             "beta2" = 0.25,
             "beta3" = 0.35,
             "phi" = 0.14,
             "beta11" = 0.16,
             "sigma_q" = 0.021,
             "sigma_xi" = 66.89)

@@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

  for(i in 1:length(ovov_sim_100)){
    dat <- data.frame(time=c(1:416), ovov_sim_100[[i]])
    dat |>
      pomp(t0=0,
           time="time",
           rprocess=discrete_time(rproc,delta.t=1/4), ## Obs time every week, but delta.t transition is actually everyday
           rinit=rinit_wwr,
           dmeasure=dmeas,
           rmeasure=rmeas,
           accumvars=c("H1","H2","H3"),
           statenames=c("S1","S2","S3",
                        "I1","I2","I3",
                        "R1","R2","R3",
                        "H1","H2","H3",
                        "q1","q2","q3"),
           paramnames = names(wwr_mle_raw)
      ) -> sir_wwr

    theta <- wwr_mle_raw

    foreach(i=1:ncores, .combine=c,
            .options.future=list(seed=998468235L)
    ) %dofuture% {
      pfilter(sir_wwr,Np=NP_MIF,save.states=T, params=theta)
    } -> pfs

@@

bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_100.rds"),{
  pal_log_lik <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    realdat <- ovov_sim_100[[i]] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
                                     c(initial_guess[1],initial_guess[2],initial_guess[3],
                                       initial_guess[4],initial_guess[5]), 
                                     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
                                     norm_par = c(0.07,initial_guess[6]), 
                                     t(prop), NP_MIF,1)
    pal_log_lik[i] <- lik_list$log_lik
  }
  pal_log_lik
})

@@

<<plot1, echo=FALSE, fig.height=4, fig.width=5.5>>=
# Create your data
time <- 1:416
ncores <- length(list_cond_lik)

@@

# Plot with ggplot
ggplot(df, aes(x = time, y = cond_loglik, group = interaction(method, factor(1:ncores)), color = method)) +
  geom_line(data = filter(df, method == "SMC"), color = "blue") +
  geom_line(data = filter(df, method == "PAL"), color = "red") +
  theme_minimal() +
  labs(title = "Conditional Log Likelihood Over Time",
       x = "Time",
       y = "Conditional Log Likelihood",
       color = "Method") +
  theme(legend.position = "bottomright")

@

>>>>

<<OvOv-model, echo=F, warning=F, message=F>>=
# measurement model 
dmeas <- Csnippet("
                  if (ISNA(cases1)) {
                  lik = (give_log) ? 0 : 1;
                  } else {
                        lik =  dbinom(cases1, H1, q1, 1) +
                        dbinom(cases2, H2, q2, 1) +
                        dbinom(cases3, H3, q3, 1);

@@

<<maximization1, echo=F, warning=F, message=F>>=
set.seed(123)
sir_box <- list(
  beta1=c(10,15),
  beta2=c(0.2,0.4),
  beta3=c(0.3,0.5),
  phi=c(0.01,0.3),
  beta11=c(0.1,0.2),
  sigma_q=c(0.001,0.1),
  sigma_xi=c(65,70),
  S10 = c(0.01,0.05),
  I10 = c(0.0001, 0.0005),
  R10 = c(0.01, 0.05),
  S20 = c(0.01,0.7),
  I20 = c(0.000001,0.00002),
  R20 = c(0.001,0.005),
  S30 = c(0.01,0.5),
  I30 = c(0.000001, 0.0001),
  R30 = c(0.0001,0.005),
  #fixed params
  gamma = c(1,1),
  delta1=c(1/(5*52),1/(5*52)),
  delta2=c(1/(55*52),1/(55*52)),
  alpha=c(1/(78.86912*52),1/(78.86912*52)),
  mu=c(0,0),
  N=c(82372825,82372825),
  omega=c(1/(1*52),1/(1*52))
)

@@

<<maximization2, echo=F, warning=F, message=F>>=
set.seed(1234)
sd_02 = 2/3

@@

<<maximization3, echo=F, warning=F, message=F>>=
set.seed(12345)
sd_03           = (2/3)^2

@@

<<pf-sim-100, echo=F, warning=F, message=F>>=
# Set WWR's MLE
wwr_mle_raw <- c(params_fixed, 
             "beta1" = 11.48,
             "beta2" = 0.25,
             "beta3" = 0.35,
             "phi" = 0.14,
             "beta11" = 0.16,
             "sigma_q" = 0.021,
             "sigma_xi" = 66.89)

@@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
  sir_wwr <- list()
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

  for(i in 1:length(ovov_sim_100)){
  data.frame(time=c(1:416), ovov_sim_100[[i]]) |>  pomp(t0=0,
           time="time",
           rprocess=discrete_time(rproc,delta.t=1/4),
           rinit=rinit_wwr,
           dmeasure=dmeas,
           rmeasure=rmeas,
           accumvars=c("H1","H2","H3"),
           statenames=c("S1","S2","S3",
                        "I1","I2","I3",
                        "R1","R2","R3",
                        "H1","H2","H3",
                        "q1","q2","q3"),
           paramnames = names(wwr_mle_raw)
      ) -> sir_wwr[[i]]

    foreach(j=1:ncores, .combine=c,
            .options.future=list(seed=998468235L)
    ) %dofuture% {
      pfilter(sir_wwr[[i]],Np=NP_MIF,save.states=T, params=wwr_mle_raw)
    } -> pfs

@@

bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_100.rds"),{
  pal_log_lik <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    pal_sim_dat <- ovov_sim_100[[i]] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,pal_sim_dat , 9, 
                                     c(initial_guess[1],initial_guess[2],initial_guess[3],
                                       initial_guess[4],initial_guess[5]), 
                                     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
                                     norm_par = c(0.07,initial_guess[6]), 
                                     t(prop), NP_MIF,1)
    pal_log_lik[i] <- lik_list$log_lik
  }
  pal_log_lik
})

@@

<<plot1, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances. The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL. The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
# Create your data
time <- 1:416
ncores <- length(list_cond_lik)

@@

# Plot with ggplot
ggplot(df, aes(x = time, y = cond_loglik, group = interaction(method, factor(1:ncores)), color = method)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Conditional Log Likelihood Over Time",
       x = "Time",
       y = "Conditional Log Likelihood",
       color = "Method") +
  scale_color_manual(values = c("SMC" = "blue", "PAL" = "red")) +
  theme(legend.position = "bottom")
@
